### 2011-11-01 paper.tex
fill order
tidied up table
tidied
Updated

\pagestyle{myheadings}
%% ** EDIT HERE **
%% Please insert a running head of 30 characters or less.  
%% Include it twice, once between each set of braces
\newcommand\titlestring{Stochastic tree-adjoining grammars for modeling retrotransposons}
\newcommand\authorstring{
Lawrence Uricchio, Ian Holmes
}

@@

% need cite, check me, and other notes to self
\newcommand\needcite{{\bf [CITE]}}
\newcommand\checkme{{\bf [CHECK]}}

@@

\newcommand\rhs[6]{\Tree[ .$\alpha$ [ .$#1$ [ .$#3$ $\epsilon$ ] [ .$#2$ [ .$#4$ $\epsilon$ ] $\beta$ [ .$#5$ $\epsilon$ ] ] [ .$#6$ $\epsilon$ ] ] ]}
\newcommand\rhsnob[5]{\Tree[ .$\alpha$ $#2$ [ .$#1$ $#3$ $\beta$ $#4$ ] $#5$ ]}
\newcommand\rhsnouvxy[2]{\Tree[ .$\alpha$ [ .$#1$ [ .$#2$ $\beta$ ] ] ]}

@@

\newcommand\rhsA{\Tree[ .$\alpha$ [ .$U$ $\epsilon$ ] [ .$c$ $v$ $\beta$ $x$ ] $y$ ]}
\newcommand\rhsB{\Tree[ .$\alpha$ $u$ [ .$c$ $v$ $\beta$ $x$ ] [ .$Y$ $\epsilon$ ] ]}
\newcommand\rhsC{\Tree[ .$\alpha$ $u$ [ .$c$ [ .$V$ $\epsilon$ ] $\beta$ $x$ ] $y$ ]}
\newcommand\rhsD{\Tree[ .$\alpha$ $u$ [ .$c$ $v$ $\beta$ [ .$X$ $\epsilon$ ] ] $y$ ]}
\newcommand\rhsE{\rhs{B}{C}{u}{v}{x}{y}}
\newcommand\rhsF{\rhsnob{C}{u}{v}{x}{y}}

@@

\subsection{A parsing algorithm}

@@

A suitable (but non-unique) Inside fill order is as follows:
\begin{itemize}
 \item For $L_1 = 0$ to $|Z|$ (increasing), \quad $i = 0$ to $|Z|-L_1$ (increasing):
 \begin{itemize}
  \item Let $l = i + L_1$
  \item For $A \in \nodelabels$ (topo-sorted, transition destinations before sources):
  \begin{itemize}
   \item For $L_2 = L_1$ to $0$ (decreasing), \quad $j = i$ to $i + L_1 - L_2$ (increasing):
   \item Let $k = j + L_2$
   \item Compute $\m(i,j,k,l,A)$ and store
  \end{itemize}
  \item Compute $\mtip(i,l,A)$ and store
 \end{itemize}
\end{itemize}

To model genomic features, it is extremely useful to augment the TAG framework with length distributions;
that is, to allow the rule weight $\weight(\rho)$ to depend on distances
such as $m-i$, $j-m$, $n-k$, $l-n$, $n-m$, $j-i$, $k-j$, $l-k$, $l-i$ and so on.

Note that the TAG framework implicitly allows some length distributions to be constructed, even without such augmentation;
for example, the ``waiting time'' in a particular state is geometrically-distributed, and by arranging a series of such states,
one can obtain other distributions, e.g. a negative binomial distribution.
One can also obtain any distribution over a random variable $x \leq N$ by using $N$ states,
or approximate any distribution to some degree of precision with a finite number of states.
However, it is often much more convenient and efficient simply to allow $\weight$ to sometimes depend directly on the inside sequence length.

@@

@@

\noindent
\begin{tabular}{ll}
Symbol & Meaning \\
\hline
$\grammar$ & Grammar \\
$\nodelabels$ & Set of node labels \\
$\terminals$ & Set of terminals \\
$\rules$ & Set of transformation rules \\
$\rulesubset{n}{A}$ & Set of transformation rules of type $n$ (see \tabref{RuleTypes}) whose LHS is $((\beta)A)\alpha$ \\
$\alpha,\beta$ & Arbitrary ordered node-labeled trees \\
$\weight$ & Rule weight function \\
$\seqweight{\alpha}{\outseq}$ & Sum, over derivations of $\outseq$ from $\alpha$, of product over derivation rule weights \\
$\epsilon$ & The empty string \\
$\terminals^\ast$ & Set of strings over $\terminals$, including the empty string \\
$\outseq$ & Output sequence \\
$|\outseq|$ & Length of $\outseq$ \\
$\outsubseq{i}{j+1}$ & Substring of $\outseq$ from $i$ to $j$ inclusive ($i$ starts at 1) \\
$\outsubseq{i}{i}$ & The empty string \\
$\match(i,j,x)$ & Indicates if rule string $x$ matches output string $\outsubseq{i}{j}$ \\
$\fmatch(i,x)$ & Indicates if rule string $x$ matches output string $\outseq$, starting at position $i$ \\
$\rmatch(j,y)$ & Indicates if rule string $y$ matches output string $\outseq$, ending before position $j$ \\
$\iumatch,\jvmatch,\kxmatch,\lymatch$ & Rule string matchers; $\match_{ns}$ matches string $s$ ending/starting at position $n$ \\
$\m(i,j,k,l,A)$ & Inside weight for $\outsubseq{i}{j}$ (left) and $\outsubseq{k}{l}$ (right) rooted at $A$ \\
$\mtip(i,l,A)$ & Inside weight for $\outsubseq{i}{l}$ rooted at $A$ \\
$B,C,U,V,X,Y$ & Labels (or terminal strings) at nodes of the Newick tree $((U,((V,\beta,W)C),X)B)\alpha$ \\
$u,v,x,y$ & Special symbols for $U,V,X,Y$ when they are terminal strings \\
\end{tabular}

>>>>

\pagestyle{myheadings}
%% ** EDIT HERE **
%% Please insert a running head of 30 characters or less.  
%% Include it twice, once between each set of braces
\newcommand\titlestring{Stochastic grammars for transposon sequences}
\newcommand\authorstring{
Lawrence Uricchio, Ian Holmes
}

@@

% need cite, check me, and other notes to self and co-authors
\newcommand\redpen[1]{{\bf \textcolor{red}{#1}} \textcolor{black}{}}
\newcommand\needcite{\redpen{\bf [CITE]}}
\newcommand\checkme{\redpen{\bf [CHECK]}}
\newcommand\writeme{\redpen{\bf [WRITE ME!]}}

@@

\newcommand\rhs[6]{\Tree[ .$\alpha$ [ .$#1$ [ .$#3$ $\epsilon$ ] [ .$#2$ [ .$#4$ $\epsilon$ ] $\beta$ [ .$#5$ $\epsilon$ ] ] [ .$#6$ $\epsilon$ ] ] ]}
\newcommand\rhsnob[5]{\Tree[ .$\alpha$ $#2$ [ .$#1$ $#3$ $\beta$ $#4$ ] $#5$ ]}
\newcommand\rhsnoepsilon[6]{\Tree[ .$\alpha$ [ .$#1$ $#3$ [ .$#2$ $#4$ $\beta$ $#5$ ] $#6$ ] ]}
\newcommand\rhsnouvxy[2]{\Tree[ .$\alpha$ [ .$#1$ [ .$#2$ $\beta$ ] ] ]}

@@

\newcommand\rhsA{\Tree[ .$\alpha$ [ .$U$ $\epsilon$ ] [ .$C$ $v$ $\beta$ $x$ ] $y$ ]}
\newcommand\rhsB{\Tree[ .$\alpha$ $u$ [ .$C$ $v$ $\beta$ $x$ ] [ .$Y$ $\epsilon$ ] ]}
\newcommand\rhsC{\Tree[ .$\alpha$ $u$ [ .$C$ [ .$V$ $\epsilon$ ] $\beta$ $x$ ] $y$ ]}
\newcommand\rhsD{\Tree[ .$\alpha$ $u$ [ .$C$ $v$ $\beta$ [ .$X$ $\epsilon$ ] ] $y$ ]}
\newcommand\rhsE{\rhsnoepsilon{B}{C}{u}{v}{x}{y}}
\newcommand\rhsF{\rhsnob{C}{u}{v}{x}{y}}

@@

\newcommand\outerlen{L_0}
\newcommand\innerlen{L_1}

To model genomic features, it is extremely useful to augment the TAG framework with length distributions;
that is, to allow the rule weight $\weight(\rho)$ to depend on distances
such as $i$, $|\outseq|-l$, $\outerlen = l-i$, $\innerlen = k-j$, $j-i$, $l-k$, $m-i$, $j-m$, $n-k$, $l-n$ or $n-m$.

Note that the TAG framework implicitly allows some length distributions to be constructed, even without such augmentation;
for example, the ``waiting time'' in a particular state is geometrically-distributed, and by arranging a series of such states,
one can obtain other distributions, e.g. a negative binomial distribution.
One can also obtain any distribution over a random variable $x \leq N$ by using $N$ states,
or approximate any distribution to some degree of precision with a finite number of states.
However, it is sometimes more convenient simply to allow $\weight$ to depend directly on $(i,j,k,l,m,n)$.

\subsection{A bottom-up (CYK) parsing algorithm}

@@

A suitable (but non-unique) Inside fill order is as follows:
\begin{itemize}
 \item For $\outerlen = 0$ to $|\outseq|$ (increasing), \quad $i = 0$ to $|\outseq|-\outerlen$ (increasing):
 \begin{itemize}
  \item Let $l = i + \outerlen$
  \item For $\innerlen = \outerlen$ to $0$ (decreasing), \quad $j = i$ to $i + \outerlen - \innerlen$ (increasing):
  \begin{itemize}
   \item Let $k = j + \innerlen$
   \item For $A \in \nodelabels$ (topo-sorted, transition destinations before sources):
   \begin{itemize}
    \item Compute $\m(i,j,k,l,A)$ and store
   \end{itemize}
  \end{itemize}
  \item For $A \in \nodelabels$ (topo-sorted, transition destinations before sources):
  \begin{itemize}
   \item Compute $\mtip(i,l,A)$ and store
  \end{itemize}
 \end{itemize}
\end{itemize}

\subsection{A top-down parsing algorithm}
\seclabel{TopDownParser}

The general idea is to try a top-down chart parser, such as the Earley parser \cite{Earley70}
(a version of which exists for TAGs \cite{kallmeyer2010parsing})
that can run very efficiently on preprocessed sequence,
where the preprocessor augments the sequence with a table of tuples of the form $(i,j,k,l,A,\m)$
and the parser restricts the iteration to derivations that use some of these $(i,j,k,l,A)$ co-ordinates.

\writeme

@@

\section{A simple DNA transposon grammar}
\seclabel{TransposonGrammar}

\writeme

@@

\noindent
\begin{tabular}{ll}
Symbol & Meaning \\
\hline
$\grammar$ & Grammar \\
$\nodelabels$ & Set of node labels \\
$\terminals$ & Set of terminals \\
$\rules$ & Set of transformation rules \\
$\rulesubset{n}{A}$ & Set of transformation rules of type $n$ (see \tabref{RuleTypes}) whose LHS is $((\beta)A)\alpha$ \\
$\alpha,\beta$ & Arbitrary ordered node-labeled trees \\
$\weight$ & Rule weight function \\
$\seqweight{\alpha}{\outseq}$ & Sum, over derivations of $\outseq$ from $\alpha$, of product over derivation rule weights \\
$\epsilon$ & The empty string \\
$\terminals^\ast$ & Set of strings over $\terminals$, including the empty string \\
$\outseq$ & Output sequence \\
$|\outseq|$ & Length of $\outseq$ \\
$i,j,k,l$ & Integer co-ordinates on $\outseq$, with $1 \leq i \leq j \leq k \leq l \leq |\outseq|$ \\
$m,n$ & Integer co-ordinates on $\outseq$, with $i \leq m \leq j$ and $k \leq n \leq l$ \\
$\outerlen, \innerlen$ & Integer lengths of $\outseq$-subsequences, with $0 \leq \innerlen = (k-j) \leq \outerlen = (l-i) \leq |\outseq|$ \\
$\outsubseq{i}{i}$ & The empty string \\
$\outsubseq{i}{j}$ & (For $j \geq i$) Substring of $\outseq$ from $i$ to $j-1$ inclusive \\
$\match(i,j,x)$ & Indicates if rule string $x$ matches output string $\outsubseq{i}{j}$ \\
$\fmatch(i,x)$ & Indicates if rule string $x$ matches output string $\outseq$, starting at position $i$ \\
$\rmatch(j,y)$ & Indicates if rule string $y$ matches output string $\outseq$, ending before position $j$ \\
$\iumatch,\jvmatch,\kxmatch,\lymatch$ & Rule string matchers; $\match_{ns}$ matches string $s$ ending/starting at position $n$ \\
$\m(i,j,k,l,A)$ & Inside weight for $\outsubseq{i}{j}$ (left) and $\outsubseq{k}{l}$ (right) rooted at $A$ \\
$\mtip(i,l,A)$ & Inside weight for $\outsubseq{i}{l}$ rooted at $A$ \\
$B,C,U,V,X,Y$ & Labels (or terminal strings) at nodes of the Newick tree $((U,((V,\beta,W)C),X)B)\alpha$ \\
$u,v,x,y$ & Special symbols for $U,V,X,Y$ when they are terminal strings \\
\end{tabular}
